---
title: "Lab 12 - Smoking during pregnancy"
author: "Ben Waggener"
date: "4/11/2025"
output: github_document
---

### Load packages and data

```{r load-packages, message=FALSE}
library(tidyverse) 
library(tidymodels)
library(openintro)
```

setting a seed for reproducibility 
```{r seed}
set.seed(04102025)
```

loading data
```{r get data}
ncbirths <- openintro::ncbirths
```



### Exercise 1

cases are individual births such that it includes information aboout the child, the mother, and the father. There are 1000 cases in this sample.

### Exercise 2

filter to keep only the white mothers

Mean weight of babies from white moms : 7.25  
```{r filter_white}
ncbirths_white <- ncbirths %>%
  filter(whitemom == "white")

summary(ncbirths_white)

obs_mean <- ncbirths_white %>%
  summarize(mean_weight = mean(weight)) 
```

### Exercise 3
the criteria for conducting simulation-based inference are satisfied. The data has a lot of observations (1000), they are randomly sampled, and these observations are independent of each other.

##Exercise 4
H0:μ=7.43 poundsH0:μ=7.43 pounds.
HA:μ≠7.43 poundsHA:μ≠7.43 pounds.

There were 0 bootstrapped samples with the null distribution wiht a mean at or lower than 7.25 (p < .05). This means that it is highly unlikely that the mean birth weight in 1995 was the same as it is now, they are significantly different. 

```{r nulldist}
sim_ncbirth_white <- ncbirths_white %>%
  specify(response = weight) %>%
  generate(reps = 100, type = "bootstrap") %>%
  calculate(stat = "mean")
summary(sim_ncbirth_white)
```
```{r shift}

obs_mean_value <- obs_mean$mean_weight

# Calculate shift value
shift_amount <- obs_mean_value - 7.43

# Shift bootstrap distribution
null_dist <- sim_ncbirth_white %>%
  mutate(stat = stat - shift_amount)
```

```{r plot}
ggplot(null_dist, mapping = aes(x = stat))+
  geom_histogram(binwidth = .01)+
  geom_vline(xintercept = 7.25, color = "red", linetype = "dashed") +
  labs(
    title = "Null Distribution",
    x = "Sample Mean Birth Weight (lbs)",
       y = "Count")+
  theme_classic()
```

```{r pvalue}
null_dist %>%
  filter(stat <= (7.25)) %>%
  summarize(p_value = n()/nrow(null_dist))
```
```{r pvalue}
p_value <- null_dist %>%
  get_p_value(obs_stat = obs_mean_value, direction = "two_sided")

p_value
```


##Exercise 5

This plot shows that smoking reduces the childs weight compared to not smoking. Also there seems to be a lot of variability in the non smoking data.
```{r boxplot}


ggplot(ncbirths, aes(x = habit, y = weight)) +
  geom_boxplot()
```

##Exercise 6
```{r cleaning}
ncbirths_clean <- ncbirths %>%
  filter(!is.na(habit)) %>%
  filter(!is.na(weight)) 
```

##Exercise 7
```{r diff}
ncbirths_clean %>%
  group_by(habit) %>%
  summarize(mean_weight = mean(weight))
```

##Exercise 8
H0:weight of smoker = weight of nonsmoker
HA: weight of smoker ≠ weight of nonsmoker

##Exercise 9


```{r mean_diff}
obs_diff_smok <- ncbirths_clean %>%
  group_by(habit) %>%
  summarize(mean_weight = mean(weight, na.rm = TRUE)) %>%
  spread(key = habit, value = mean_weight) %>%  # Spread the data into two columns (smoker vs nonsmoker)
  mutate(diff = nonsmoker - smoker) 
```

```{r null_dist}
sim_ncbirth_smok <- ncbirths_clean %>%
  specify(response = weight, x = habit) %>%   # Specify the response and predictor variable
  generate(reps = 100, type = "bootstrap") %>% # Generate 1000 bootstrap samples
  calculate(stat = "mean") 
```